---
title: "Raphanus_specimen_map"
author: "Julia Harenčár"
date: "07/05/2023"
execute:
  echo: true #will keep code chunks
  warning: false #removes warning functions
format: 
  html: 
    self-contained: true #makes one html file to render that is sharable
    page-layout: full #uses the full page
    code-fold: true #lets user fold up codes
    code-tools: true #makes an option in the upper left for users to reveal/hide all code and copy source code
    #makes long lines of code rap
    code-overflow: wrap
    #makes table of contents
    toc: true
    #sets location of table of contents
    toc-location: left
#set theme settings. in this case user can switch between light and dark mode
theme:
  #give light option
  light: flatly
  #give dark option
  dark: darkly
---

# Raphanus specimen maps

Creating maps of *Raphanus* specimen locations and collection years.\
\

```{r load libraries}
# load libraries
library(tidyverse)
library(ggplot2)
theme_set(theme_bw())
library(maps)
library(ggmap)
library(grid)
library(patchwork)
#library(plotly)
library(htmlwidgets)
library(raster)
library(sf)
library(leaflet)
library(leaflet.extras)
library(RColorBrewer)

library(ade4) # PCA computation

setwd("/Users/juliaharencar/Documents/Github/Raphanus_specimen_map")
```

### Specimen list building

```{r}
# import full dataset and list of specimen IDs selected based on map and age
CA_rad_spec <- read.csv("Raphanus_specimens_all_2024.csv", header = T)
samp_list <- read.csv("Raphanus_specimen_sampling_incomp.csv", header = T)

# add specimen location gps coordinates from Ntalie's hand curation
fulldata<- read.csv("/Users/juliaharencar/Documents/Github/Raphanus_Phenology/Output_Files/completed_specimen_data.csv", header = T)
CA_rad_spec <- CA_rad_spec %>%
  rows_update(
    fulldata %>% dplyr::select(id, specificEpithet, institutionCode, decimalLatitude, decimalLongitude, year),
    by = "id",
    unmatched = "ignore"
  )

# Extract the IDs from the list of specimens to sample
selected <- samp_list$id

# Subset the complete dataframe based on the IDs
samp_list_w.data <- CA_rad_spec %>%
  filter(id %in% selected)

# print new sampling list with all metadata
library(writexl)
write_xlsx(samp_list_w.data, "Raphanus_sampling_specimens.xlsx")

# quick histogram distibution of years
hist(samp_list_w.data$year, breaks=60)

ggplot(samp_list_w.data, aes(x=year)) +
  geom_histogram(col = "black", fill = "darkolivegreen3") +
  scale_x_continuous(breaks = round(seq(1850, max(samp_list_w.data$year), by = 10),1)) +
  scale_y_continuous(breaks = seq(10,70, by=10))

#   geom_point() +
#   labs(title = "Pos
```

### newest map with sampling

```{r}
# new map colored by year to check sample distribution/make sure I didn't miss anything key 
# NOTE: might need to run some of below lines for mapping stuff
colors <- colorNumeric("Spectral", domain = CA_rad_spec$year)

CA_rad_spec <- CA_rad_spec %>% filter(institutionCode=="DAV")

full_specimen_map_w_sampling <- leaflet(CA_rad_spec) %>%
  addTiles() %>%  # Add OpenStreetMap tiles
  # addRasterImage(precip_raster_CA,
  #                opacity = 0.7,  # Adjust opacity as needed
  #                colors = brewer.pal(9,'YlGnBu')) %>%  # Use Spectral color palette for raster
  addCircleMarkers(
    lat = ~decimalLatitude, 
    lng = ~decimalLongitude, 
    radius = 4,  # Size of the markers
    color = ~colors(year),
    stroke = FALSE, fillOpacity = 1,  # Set fill opacity to 1 for fully opaque markers
    popup = ~paste("Scientific Name: ", specificEpithet, "<br>Year: ", year, "Herbarium:", institutionCode, "ID:", id),  # Popup with info
    label = ~paste(specificEpithet, institutionCode, year, id),  # Add data as labels for searching
    group = "allPoints"
  ) %>%
  # Add selected points with a dark border
  addCircleMarkers(
    data = CA_rad_spec[CA_rad_spec$id %in% selected, ],
    lat = ~decimalLatitude, 
    lng = ~decimalLongitude, 
    radius = 4,  # Same size as the other points
    color = "black",  # Border color
    stroke = TRUE, weight = 3,  # Enable stroke and set border thickness
    fillColor = ~colors(year),  # Keep the fill color the same
    fillOpacity = 1,  # Fully opaque fill
    popup = ~paste("Scientific Name: ", specificEpithet, "<br>Year: ", year, "Herbarium:", institutionCode, "ID:", id),  # Popup with info
    label = ~paste(specificEpithet, institutionCode, year, id),  # Add id as a label for searching
    group = "selectedPoints"  # Separate group for selected points
  ) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6) %>%  # Set the view over California
  addLegend(
    position = "bottomright", 
    pal = colorNumeric("Spectral", CA_rad_spec$year), 
    values = ~year,
    title = "Year",
    opacity = 1
  ) %>%
  # Add a search bar to find points by ID
  addSearchFeatures(
    targetGroups = "allPoints",  # Search within all markers
    options = searchFeaturesOptions(
      zoom = 10,  # Zoom level when a match is found
      openPopup = TRUE,  # Open the popup of the matching point
      textPlaceholder = "Search map..."  # Placeholder text in the search bar
 )
  )

saveWidget(full_specimen_map_w_sampling, "bold_selected_CA_specimen_color_by_year_new_DAV.html")
```

##Collected specimen maps
```{r}
# import full dataset and list of sampled specimen IDs
CA_rad_spec <- read.csv("Raphanus_specimens_all_2025.csv", header = T)
samp_list <- read.csv("Herbarium_tissue_collected_tmp_251114.csv", header = T)

# add specimen location gps coordinates from Natalie and Juliette's hand curation
Nat_coords<- read.csv("/Users/juliaharencar/Documents/Github/Raphanus_Phenology/Output_Files/completed_specimen_data.csv", header = T)
more_coords <- read.csv("CA_rad_spec_1915to40_missing.coords.csv", header = T)
more_coords$decimalLatitude <- as.numeric(more_coords$decimalLatitude)
yet_more <- read.csv("CA_rad_spec_sampled.to.251114_missing.coords_w_some_filled.csv", header = T)
yet_more$decimalLatitude <- as.numeric(yet_more$decimalLatitude)

CA_rad_spec <- CA_rad_spec %>%
  rows_update(
    Nat_coords %>% dplyr::select(id, decimalLatitude, decimalLongitude),
    by = "id",
    unmatched = "ignore"
  ) %>% 
  rows_update(
    more_coords %>% dplyr::select(id, decimalLatitude, decimalLongitude),
    by = "id",
    unmatched = "ignore"
  ) %>% 
  rows_update(
    yet_more %>% dplyr::select(id, decimalLatitude, decimalLongitude),
    by = "id",
    unmatched = "ignore"
  )

# # get list of specimens without GPS data:
# missing.coords <- CA_rad_spec %>% filter(is.na(decimalLatitude))

# Extract the IDs from the list of sampled specimens
selected <- samp_list$id

# Subset the complete dataframe based on the IDs
samp_list_w.data <- CA_rad_spec %>%
  filter(id %in% selected)

# histogram: distibution of years 
ggplot(samp_list_w.data, aes(x = year)) +
  geom_histogram(binwidth = 5, col = "black", fill = "darkolivegreen3") +
  scale_x_continuous(breaks = round(seq(1850, max(samp_list_w.data$year), by = 5), 1)) +
  scale_y_continuous(breaks = seq(5, 70, by = 5)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

### new map colored by year to check sample distribution/make sure I didn't miss anything key 
# NOTE: might need to run some of below lines for mapping stuff

# if I want to limit it to just what I sampled
# CA_rad_spec <- CA_rad_spec %>%
#  filter(id %in% selected)

# Function to make small equilateral triangle icons centered on the point
make_triangle_uri <- function(hex_color, size = 10, stroke = "black", stroke_width = 2) {
  col_enc <- gsub("#", "%23", hex_color)
  sprintf(
    "data:image/svg+xml;utf8,
    <svg xmlns='http://www.w3.org/2000/svg' width='%d' height='%d' viewBox='0 0 20 20'>
      <polygon points='10,3 18,17 2,17' fill='%s' stroke='%s' stroke-width='%.2f'/>
    </svg>",
    size, size, col_enc, stroke, stroke_width
  )
}

# Prepare data and color palette
selected_df <- CA_rad_spec %>% filter(id %in% selected)
colors <- colorNumeric("Spectral", CA_rad_spec$year)

# Generate triangle icons for selected points
icon_size <- 10
icon_urls <- vapply(colors(selected_df$year),
                    FUN = make_triangle_uri,
                    FUN.VALUE = character(1),
                    size = icon_size)

tri_icons <- icons(
  iconUrl = icon_urls,
  iconWidth = icon_size,
  iconHeight = icon_size,
  iconAnchorX = round(icon_size / 2),
  iconAnchorY = round(icon_size / 2)
)

# Build leaflet map
full_specimen_map_w_sampling <- leaflet(CA_rad_spec) %>%
  addTiles() %>%
  
  # Base layer: all points
  addCircleMarkers(
    lat = ~decimalLatitude, 
    lng = ~decimalLongitude, 
    radius = 4,
    color = ~colors(year),
    stroke = FALSE,
    fillOpacity = 1,
    popup = ~paste("Scientific Name:", specificEpithet,
                   "<br>Year:", year,
                   "<br>Herbarium:", institutionCode,
                   "<br>ID:", id),
    label = ~paste(specificEpithet, institutionCode, year, id),
    group = "allPoints"
  ) %>%
  
  # Selected subset as triangles
  addMarkers(
    data = selected_df,
    lat = ~decimalLatitude,
    lng = ~decimalLongitude,
    icon = tri_icons,
    popup = ~paste("Scientific Name:", specificEpithet,
                   "<br>Year:", year,
                   "<br>Herbarium:", institutionCode,
                   "<br>ID:", id),
    label = ~paste(specificEpithet, institutionCode, year, id),
    group = "selectedPoints",
    options = markerOptions(riseOnHover = TRUE)
  ) %>%
  
  # Legend, search, and view
  setView(lng = -119.4179, lat = 36.7783, zoom = 6) %>%
  addLegend(
    position = "bottomright", 
    pal = colors, 
    values = ~year,
    title = "Year",
    opacity = 1
  ) %>%
  addSearchFeatures(
    targetGroups = "allPoints",
    options = searchFeaturesOptions(
      zoom = 10,
      openPopup = TRUE,
      textPlaceholder = "Search map..."
    )
  )

saveWidget(full_specimen_map_w_sampling, "triangle_sampled_CA_specimen_color_by_year_251125.html")
```
### climate PC1 color on sampled map
#### clim PCA

```{r}
# import data
CA_rad_spec <- read.csv("CCH2_all_2024.csv", header = T)
post_2000 <- CA_rad_spec %>% dplyr::select(specificEpithet, year, decimalLatitude, decimalLongitude, county, institutionCode, id) %>% filter(year >= 2000)
# s2015_e2019 <- CA_rad_spec %>% dplyr::select(specificEpithet, year, decimalLatitude, decimalLongitude, county) %>% filter(year >= 2015 & year <= 2019)

# order by longitude
post_2000 <- post_2000 %>% arrange(desc(decimalLatitude))
# s2015_e2019 <- s2015_e2019 %>% arrange(desc(decimalLatitude))

# import climate data:
# DD5 = growing degree-days
# MAP = mean annual precip
# MAT = mean annual temp
# MCMT = mean coldest month temperature (°C)
# MWMT = mean warmest month temperature (°C)
# SHM = summer heat-moisture index ((MWMT)/(MSP/1000))
tiff_files <- list.files(path=".", pattern = "*.tif")
rasters <- lapply(tiff_files, raster::raster)

# precip_raster <- raster::raster("MAP.tif")

# Check the projection 
#crs(precip_raster) # '+proj=longlat +datum=WGS84 +no_defs"

## extract data just for specific coordinates

# get coordinates
coordinates <- post_2000 %>% dplyr::select(decimalLongitude,decimalLatitude)
# coordinates <- s2015_e2019 %>% dplyr::select(decimalLongitude,decimalLatitude)
names(coordinates) <- c("longitude", "latitude")

# Define empty data frame to store results
climate_compiled <- data.frame(longitude = coordinates[,1], latitude = coordinates[,2])

# Loop through each raster and extract data
for (i in 1:length(rasters)) {
  current_raster <- rasters[[i]]
  current_data <- extract(current_raster, SpatialPoints(coordinates, proj4string = CRS("+proj=longlat +datum=WGS84 +no_defs ")), sp = T)
  current_df <- as.data.frame(current_data)
  # Add new column with data from current raster
  climate_compiled[, paste0("raster_", i)] <- current_df[,1]
}

names(climate_compiled) <- c("longitude", "latitude", "DD5", "MAP", "MAT", "MCMT", "MWMT", "SHM")

# generate Climate PCA 
# # add identifier for each specimen 
# climate_compiled <- climate_compiled %>%
#   mutate(specimen_no = row_number())

# remove coordinates
climate_data <- climate_compiled[, -c(1, 2)]

# calculate principal components
clim_pca <- prcomp(climate_data, center = TRUE, scale = TRUE)

# # reverse signs (eigenvectors in R point in negative direction by default)
# clim_pca$rotation <- -1*clim_pca$rotation
# clim_pca$x <- -1*clim_pca$x

# display principal components
clim_pca$rotation

# Basic biplot
biplot(clim_pca, scale = 0)

# extract PCA scores for plotting
pca_scores <- as.data.frame(clim_pca$x)

# add location info
#pca_scores <- bind_cols(post_2000 %>% dplyr::select(county), pca_scores)
pca_scores_post_2000 <- bind_cols(post_2000 %>% dplyr::select(specificEpithet, year, decimalLatitude, decimalLongitude, county, institutionCode, id), pca_scores)
                        
# Calculate the variance explained by each principal component
per_variance_explained <- (clim_pca$sdev^2 / sum(clim_pca$sdev^2))*100
# Create a data frame for the variance explained
variance_df <- data.frame(
  PC = paste0("PC", 1:length(per_variance_explained)),
  Variance_Explained = per_variance_explained,
  Cumulative_Variance = cumsum(per_variance_explained)
)

## plot with ggplot
# # create PC loading vectors for plot
# loadings <- clim_pca$rotation
# # calculate arrow coordinates
# start_x <- rep(0, ncol(loadings))
# start_y <- rep(0, ncol(loadings))
# end_x <- loadings[, 1] * max(abs(clim_pca$x[, 1]))
# end_y <- loadings[, 2] * max(abs(clim_pca$x[, 2]))

## plot no loadings
# ggplot(pca_scores, aes(x=PC1, y=PC2, colour = county)) +
#   geom_point() +
#   labs(title = "Post 2000 specimens", x = "PC1 (66.6%)", y = "PC2 (24.5%)")
ggplot(pca_scores, aes(x=PC1, y=PC2, colour = county)) +
  geom_point() +
  labs(title = "Specimens from 2015-2019", x = "PC1 (72.6%)", y = "PC2 (17.6%)")
  
ggsave("Climate_PCA_2015-2019.png", width = 20, height = 10, units = "cm")

## biplot
biplot(clim_pca, cex = c(0.5,1))
ggbiplot(clim_pca)
ggsave("Climate_biplot_2015-2019.png", width = 15, height = 15, units = "cm")
```

#### PC1 color map

```{r}
# import data
#CA_rad_spec <- read.csv("CCH2_all_2024.csv", header = T)

#CA_rad_spec_2015_2024 <- CA_rad_spec %>% filter(year >= 2015 & year <= 2024)

# Load the GeoTIFF file
precip_raster <- raster("MAP.tif")

# Load the California shapefile
california_shape <- st_read("./ca_state/CA_State.shp")

# Reproject raster to match the CRS of the shapefile
precip_raster_reproj <- projectRaster(precip_raster, crs = st_crs(california_shape)$proj4string)
# Crop the raster using the California shapefile
precip_raster_CA <- mask(crop(precip_raster_reproj, extent(california_shape)), california_shape)

# assign coordinate system with right syntax (this is the system the raster is in)
precip_raster_CA <- projectRaster(precip_raster_CA, crs = "+proj=longlat +datum=WGS84")

# not working and apparently not needed
# convert for leaflet
#leaflet::projectRasterForLeaflet(precip_raster_CA)

# Create a custom reversed palette function
reversed_spectral <- function(n) {
  original_palette <- brewer.pal(n, "Spectral")
  rev(original_palette)
}

reversed_PuOr <- function(n) {
  original_palette <- brewer.pal(n, "PuOr")
  rev(original_palette)
}

custom_palette <- colorRampPalette(c("#2166AC", "#4393C3", "#92C5DE", "#F4A582", "#D6604D", "#B2182B"))

pal <- colorNumeric(palette = custom_palette(100), domain = pca_scores_post_2000$PC1)

specimen_map_leaflet <- leaflet(pca_scores_post_2000 ) %>%
  addTiles() %>%  # Add OpenStreetMap tiles
  addCircleMarkers(
    lat = ~decimalLatitude, 
    lng = ~decimalLongitude, 
    radius = 7,  # Size of the markers
    color = ~pal(PC1),
    stroke = FALSE, fillOpacity = 1,  # Set fill opacity to 1 for fully opaque markers
    popup = ~paste("Scientific Name: ", specificEpithet, "<br>Year: ", year, "Herbarium:", institutionCode, "ID:", id),  # Popup with info
    label = ~paste(specificEpithet, institutionCode, id),  # add id as a label for searching
    group = "allPoints"       # Assign a generic group name
  ) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6) %>%  # Set the view over California
  addLegend(
    position = "bottomright", 
    pal = colorNumeric(reversed_spectral(9), CA_rad_spec$PC1), 
    values = ~PC1,
    title = "PC1",
    opacity = 1
  ) %>%
  # Add a search bar to find points by ID
  addSearchFeatures(
    targetGroups = "allPoints",  # Search within all markers
    options = searchFeaturesOptions(
      zoom = 10,  # Zoom level when a match is found
      openPopup = TRUE,  # Open the popup of the matching point
      textPlaceholder = "Search map..."  # Placeholder text in the search bar
    )
  )

saveWidget(specimen_map_leaflet, "interactive_CA_specimen_color_by_PC1_post_2000.html")
```
### 15-40/00-25 subset maps
```{r}
### Plot only post 2000s and 1915-1940 specimens to design first round of sampling
CA_rad_spec_1915to40_post2000 <- CA_rad_spec %>%
  filter((year >= 1915 & year <= 1940) | year > 2000)

# print out full list to play with in excel, including col showing whether it has been sampled
CA_rad_spec_1915to40_post2000 %>%
  mutate(sampled = if_else(id %in% selected, "y", "n")) %>%
  write.csv(file = "CA_rad_spec_1915to40_post2000.csv", row.names = FALSE)

# list those from the same time period missing lat and long to know what I could add with some location research:
CA_rad_spec_1915to40_post2000_missing.coords <- CA_rad_spec_1915to40_post2000 %>%
  filter(is.na(decimalLatitude))
# save list to see about getting location data:
write.csv(CA_rad_spec_1915to40_post2000_missing.coords, file = "CA_rad_spec_1915to40_post2000_missing.coords.csv", row.names = FALSE)

# Make a categorical variable for time period
CA_rad_spec_1915to40_post2000 <- CA_rad_spec_1915to40_post2000 %>%
  mutate(year = as.numeric(year)) %>%
  mutate(period = case_when(
    year >= 1915 & year <= 1940 ~ "1915–1940",
    year > 2000 ~ "Post-2000",
    TRUE ~ NA_character_
  ))

# Explicitly coerce the domain to a plain character vector
domain_vec <- as.character(CA_rad_spec_1915to40_post2000)

# Define a simple two-color palette (warm/cool)
period_pal <- colorFactor(
  palette = c("red", "#B66DFF"),
  domain = domain_vec,
  levels = c("1915–1940", "Post-2000")  # order/consistency
)
  
# make small, equilateral triangle centered at the coordinate
make_triangle_uri <- function(hex_color, size = 10, stroke = "black", stroke_width = 2) {
  col_enc <- gsub("#", "%23", hex_color)
  
  # Equilateral triangle centered roughly at (10,10)
  svg <- sprintf(
    "data:image/svg+xml;utf8,<svg xmlns='http://www.w3.org/2000/svg' width='%d' height='%d' viewBox='0 0 20 20'>
      <polygon points='10,3 18,17 2,17' fill='%s' stroke='%s' stroke-width='%.2f'/>
    </svg>",
    size, size, col_enc, stroke, stroke_width
  )
  svg
}

icon_size <- 10  # triangle size

sel_df <- CA_rad_spec_1915to40_post2000 %>% filter(id %in% selected)

icon_urls <- vapply(period_pal(sel_df$period),
                    FUN = make_triangle_uri,
                    FUN.VALUE = character(1),
                    size = icon_size)

# Center anchor for alignment
tri_icons <- icons(
  iconUrl = icon_urls,
  iconWidth = icon_size,
  iconHeight = icon_size,
  iconAnchorX = round(icon_size / 2),
  iconAnchorY = round(icon_size / 2)
)

# Plot the map
specimen_map_w_sampling_1915to40.post2000 <- leaflet(CA_rad_spec_1915to40_post2000) %>%
  addTiles() %>%
  
  addCircleMarkers(
    lat = ~decimalLatitude,
    lng = ~decimalLongitude,
    radius = 4,
    fillColor = ~period_pal(period),
    color = ~period_pal(period),
    stroke = FALSE,
    fillOpacity = 0.9,
    popup = ~paste("Scientific Name:", specificEpithet, "<br>Year:", year,
                   "<br>Herbarium:", institutionCode, "<br>ID:", id),
    label = ~paste(specificEpithet, institutionCode, year, id),
    group = "allPoints"
  ) %>%
  
  addMarkers(
    data = sel_df,
    lat = ~decimalLatitude,
    lng = ~decimalLongitude,
    icon = tri_icons,
    popup = ~paste("Scientific Name:", specificEpithet, "<br>Year:", year,
                   "<br>Herbarium:", institutionCode, "<br>ID:", id),
    label = ~paste(specificEpithet, institutionCode, year, id),
    group = "selectedPoints",
    options = markerOptions(riseOnHover = TRUE)
  ) %>%
  
  addLegend(
    position = "bottomright",
    pal = period_pal,
    values = ~period,
    title = "Collection Period",
    opacity = 1
  ) %>%
  
  addSearchFeatures(
    targetGroups = "allPoints",
    options = searchFeaturesOptions(
      zoom = 10, openPopup = TRUE,
      textPlaceholder = "Search map..."
    )
  )

saveWidget(specimen_map_w_sampling_1915to40.post2000, "triangle_sampled_CA_specimen_color_by_year_251024_subset15to40.post2000.html")
```
### extracted over sampled maps
```{r}
# import full dataset and list of sampled specimen IDs
CA_rad_spec <- read.csv("Raphanus_specimens_all_2025.csv", header = T)
samp_list <- read.csv("Herbarium_tissue_collected_tmp_251114.csv", header = T)
# import lists of extracted specimens
extracted <- read.csv("Raphanus_lab_database_tmp_251106.csv", header = T)
colnames(extracted)[2] <- "id"

# add specimen location gps coordinates from Natalie and Juliette's hand curation
Nat_coords<- read.csv("/Users/juliaharencar/Documents/Github/Raphanus_Phenology/Output_Files/completed_specimen_data.csv", header = T)
more_coords <- read.csv("CA_rad_spec_1915to40_missing.coords.csv", header = T)
more_coords$decimalLatitude <- as.numeric(more_coords$decimalLatitude)
yet_more <- read.csv("CA_rad_spec_sampled.to.251114_missing.coords_w_some_filled.csv", header = T)
yet_more$decimalLatitude <- as.numeric(yet_more$decimalLatitude)

CA_rad_spec <- CA_rad_spec %>%
  rows_update(
    Nat_coords %>% dplyr::select(id, decimalLatitude, decimalLongitude),
    by = "id",
    unmatched = "ignore"
  ) %>% 
  rows_update(
    more_coords %>% dplyr::select(id, decimalLatitude, decimalLongitude),
    by = "id",
    unmatched = "ignore"
  ) %>% 
  rows_update(
    yet_more %>% dplyr::select(id, decimalLatitude, decimalLongitude),
    by = "id",
    unmatched = "ignore"
  )

# Extract the IDs from the list of sampled specimens
selected <- samp_list$id

# Subset the complete dataframe based on the IDs
samp_list_w.data <- CA_rad_spec %>%
  filter(id %in% selected)

# Extract IDs from list of extracted specimens
extracted <- extracted$id 

# ### Plot only post 2000s and 1915-1940 specimens to design first round of sampling
# samp_list_w.data_1915to40_post2000 <- samp_list_w.data %>%
#   filter((year >= 1915 & year <= 1940) | year > 2000)

# # print out full list to play with in excel, including col showing whether it has been sampled
# samp_list_w.data_1915to40_post2000 %>%
#   mutate(sampled = if_else(id %in% selected, "y", "n")) %>%
#   write.csv(file = "samp_list_w.data_1915to40_post2000.csv", row.names = FALSE)
# 
# # list those from the same time period missing lat and long to know what I could add with some location research:
# samp_list_w.data_1915to40_post2000_missing.coords <- samp_list_w.data_1915to40_post2000 %>%
#   filter(is.na(decimalLatitude))
# # save list to see about getting location data:
# write.csv(samp_list_w.data_1915to40_post2000_missing.coords, file = "sampled_1915to40_post2000_missing.coords_251106.csv", row.names = FALSE)

# # Make a categorical variable for time period
# samp_list_w.data_1915to40_post2000 <- samp_list_w.data_1915to40_post2000 %>%
#   mutate(year = as.numeric(year)) %>%
#   mutate(period = case_when(
#     year >= 1915 & year <= 1940 ~ "1915–1940",
#     year > 2000 ~ "Post-2000",
#     TRUE ~ NA_character_
#   ))
# 
# # Explicitly coerce the domain to a plain character vector
# domain_vec <- as.character(samp_list_w.data_1915to40_post2000)
# 
# # Define a simple two-color palette (warm/cool)
# period_pal <- colorFactor(
#   palette = c("red", "#B66DFF"),
#   domain = domain_vec,
#   levels = c("1915–1940", "Post-2000")  # order/consistency
# )
#   
# # make small, equilateral triangle centered at the coordinate
# make_triangle_uri <- function(hex_color, size = 10, stroke = "black", stroke_width = 2) {
#   col_enc <- gsub("#", "%23", hex_color)
#   
#   # Equilateral triangle centered roughly at (10,10)
#   svg <- sprintf(
#     "data:image/svg+xml;utf8,<svg xmlns='http://www.w3.org/2000/svg' width='%d' height='%d' viewBox='0 0 20 20'>
#       <polygon points='10,3 18,17 2,17' fill='%s' stroke='%s' stroke-width='%.2f'/>
#     </svg>",
#     size, size, col_enc, stroke, stroke_width
#   )
#   svg
# }
# 
# icon_size <- 10  # triangle size
# 
# ext_df <- samp_list_w.data_1915to40_post2000 %>%
#   filter(id %in% extracted) %>%
#   mutate(
#     specificEpithet = as.character(specificEpithet),
#     institutionCode = as.character(institutionCode),
#     id = as.character(id),
#     year = as.character(year)
#   )
# 
# icon_urls <- vapply(period_pal(ext_df$period),
#                     FUN = make_triangle_uri,
#                     FUN.VALUE = character(1),
#                     size = icon_size)
# 
# # Center anchor for alignment
# tri_icons <- icons(
#   iconUrl = icon_urls,
#   iconWidth = icon_size,
#   iconHeight = icon_size,
#   iconAnchorX = round(icon_size / 2),
#   iconAnchorY = round(icon_size / 2)
# )
# 
# # convert label cols to character
# samp_list_w.data_1915to40_post2000$id <- as.character(samp_list_w.data_1915to40_post2000$id)
# samp_list_w.data_1915to40_post2000$year <- as.character(samp_list_w.data_1915to40_post2000$year)
# # Plot the map
# specimen_map_w_extractions_1915to40.post2000 <- leaflet(samp_list_w.data_1915to40_post2000) %>%
#   addTiles() %>%
#   
#   addCircleMarkers(
#     lat = ~decimalLatitude,
#     lng = ~decimalLongitude,
#     radius = 4,
#     fillColor = ~period_pal(period),
#     color = ~period_pal(period),
#     stroke = FALSE,
#     fillOpacity = 0.9,
#     popup = ~paste("Scientific Name:", specificEpithet, "<br>Year:", year,
#                    "<br>Herbarium:", institutionCode, "<br>ID:", id),
#     label = ~paste(specificEpithet, institutionCode, year, id),
#     group = "allPoints"
#   ) %>%
#   
#   addMarkers(
#     data = ext_df,
#     lat = ~decimalLatitude,
#     lng = ~decimalLongitude,
#     icon = tri_icons,
#     popup = ~paste("Scientific Name:", specificEpithet, "<br>Year:", year,
#                    "<br>Herbarium:", institutionCode, "<br>ID:", id),
#     label = ~paste(specificEpithet, institutionCode, year, id),
#     group = "extractedPoints",
#     options = markerOptions(riseOnHover = TRUE)
#   ) %>%
#   
#   addLegend(
#     position = "bottomright",
#     pal = period_pal,
#     values = ~period,
#     title = "Collection Period",
#     opacity = 1
#   ) %>%
#   
#   addSearchFeatures(
#     targetGroups = "allPoints",
#     options = searchFeaturesOptions(
#       zoom = 10, openPopup = TRUE,
#       textPlaceholder = "Search map..."
#     )
#   )
# 
# saveWidget(specimen_map_w_extractions_1915to40.post2000, "triangle_extracted_CA_specimen_color_by_subset15to40.post2000_251106.html")
```

### Distance based specimen selection
```{r}
library(geosphere)
CA_rad_spec <- read.csv("Raphanus_specimens_all_2024.csv", header = T)
samp_list <- read.csv("Herbarium_tissue_collected_tmp_251022.csv", header = T)
# add specimen location gps coordinates from Ntalie's hand curation
fulldata<- read.csv("/Users/juliaharencar/Documents/Github/Raphanus_Phenology/Output_Files/completed_specimen_data.csv", header = T)
CA_rad_spec <- CA_rad_spec %>%
  rows_update(
    fulldata %>% dplyr::select(id, decimalLatitude, decimalLongitude),
    by = "id",
    unmatched = "ignore"
  )
# make separate dataframes for 20-40s and post 2000 specimens
old <- CA_rad_spec %>%
  filter(year >= 1920 & year <= 1940) %>% 
  dplyr::select(id, catalogNumber, institutionCode, specificEpithet, year, month, county, decimalLatitude, decimalLongitude)
         
new <- CA_rad_spec %>%
  filter(year > 2000) %>% 
  dplyr::select(id, catalogNumber, institutionCode, specificEpithet, year, month, county, decimalLatitude, decimalLongitude)

# calculate physical distance between all old and new points 
pairs <- expand.grid(old_id = old$id, new_id = new$id, stringsAsFactors = FALSE) %>%
  left_join(old %>% rename(old_lat = decimalLatitude, old_long = decimalLongitude), by = c("old_id" = "id")) %>%
  left_join(new %>% rename(new_lat = decimalLatitude, new_long = decimalLongitude), by = c("new_id" = "id")) %>%
  mutate(distance_km = distGeo(cbind(old_long, old_lat), cbind(new_long, new_lat)) / 1000)

# make result dataframe with rows for all pairs where the distance is less than 20km (to start, change distance to get enough sampling)
close_pairs <- pairs %>%
  filter(distance_km < 25)

# list number of samples in each group (old and recent) using unique()
cat("Unique old samples:", length(unique(close_pairs$old_id)), "\n")
cat("Unique new samples:", length(unique(close_pairs$new_id)), "\n")

# plot and bold outline points selected by the distance calculations above
old <- old %>%
  mutate(selected = if_else(id %in% close_pairs$old_id, "y", "n"))

new <- new %>%
  mutate(selected = if_else(id %in% close_pairs$new_id, "y", "n"))

ggplot() +
  geom_point(data = old, aes(x = decimalLongitude, y = decimalLatitude, color = "old", alpha = 0.6)) +
  geom_point(data = new, aes(x = decimalLongitude, y = decimalLatitude, color = "new", alpha = 0.6)) +
  geom_point(data = old %>% filter(selected == "y"), aes(x = decimalLongitude, y = decimalLatitude),
             color = "black", shape = 21, size = 3, stroke = 1.2) +
  geom_point(data = new %>% filter(selected == "y"), aes(x = decimalLongitude, y = decimalLatitude),
             color = "black", shape = 21, size = 3, stroke = 1.2) +
  scale_color_manual(values = c("old" = "steelblue", "new" = "firebrick")) +
  theme_minimal() +
  labs(title = "Old (1920–1940) vs New (>2000) Specimens",
       color = "Group", x = "Longitude", y = "Latitude")

# make df of the specimens included from both old and new cohorts, with all cols from oridinal old/new dfs 
# Get IDs of matched specimens
matched_old_ids <- unique(close_pairs$old_id)
matched_new_ids <- unique(close_pairs$new_id)

# Find nearest match for each old specimen
nearest_old <- close_pairs %>%
  group_by(old_id) %>%
  slice_min(distance_km, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(id = old_id, nearest_id = new_id, nearest_km = distance_km)

# Find nearest match for each new specimen
nearest_new <- close_pairs %>%
  group_by(new_id) %>%
  slice_min(distance_km, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(id = new_id, nearest_id = old_id, nearest_km = distance_km)

# Subset the original old/new data for only matched specimens
old_matched <- old %>%
  filter(id %in% matched_old_ids) %>%
  left_join(nearest_old, by = "id") %>%
  mutate(cohort = "old")

new_matched <- new %>%
  filter(id %in% matched_new_ids) %>%
  left_join(nearest_new, by = "id") %>%
  mutate(cohort = "new")

# Combine into one dataframe and add "sampled" columns
matched_specimens <- bind_rows(old_matched, new_matched) %>%
  mutate(
    sampled = if_else(id %in% samp_list$id, "y", "n"),
    nearest_sampled = if_else(nearest_id %in% samp_list$id, "y", "n")
  )

# Write to CSV
write.csv(matched_specimens, "CA_rad_spec_matched_25km.csv", row.names = FALSE)
```


## Climate PCA

Plotting post-2000 specimens on climate PCA to get understanding of climatic distribution

```{r}

## PICKUP
# see about data by year/think about automating import of year (and Month?? figure out if collection year is right year to associate, maybe it is previous year... depends on what month starts observation year) specific climate data for 

# import data
CA_rad_spec <- read.csv("CCH2_all_2024.csv", header = T)
post_2000 <- CA_rad_spec %>% dplyr::select(specificEpithet, year, decimalLatitude, decimalLongitude, county, institutionCode, id) %>% filter(year >= 2000)
# s2015_e2019 <- CA_rad_spec %>% dplyr::select(specificEpithet, year, decimalLatitude, decimalLongitude, county) %>% filter(year >= 2015 & year <= 2019)

# order by longitude
post_2000 <- post_2000 %>% arrange(desc(decimalLatitude))
# s2015_e2019 <- s2015_e2019 %>% arrange(desc(decimalLatitude))

# import climate data:
# DD5 = growing degree-days
# MAP = mean annual precip
# MAT = mean annual temp
# MCMT = mean coldest month temperature (°C)
# MWMT = mean warmest month temperature (°C)
# SHM = summer heat-moisture index ((MWMT)/(MSP/1000))
tiff_files <- list.files(path=".", pattern = "*.tif")
rasters <- lapply(tiff_files, raster::raster)

# precip_raster <- raster::raster("MAP.tif")

# Check the projection 
#crs(precip_raster) # '+proj=longlat +datum=WGS84 +no_defs"

## extract data just for specific coordinates

# get coordinates
coordinates <- post_2000 %>% dplyr::select(decimalLongitude,decimalLatitude)
# coordinates <- s2015_e2019 %>% dplyr::select(decimalLongitude,decimalLatitude)
names(coordinates) <- c("longitude", "latitude")

# Define empty data frame to store results
climate_compiled <- data.frame(longitude = coordinates[,1], latitude = coordinates[,2])

# Loop through each raster and extract data
for (i in 1:length(rasters)) {
  current_raster <- rasters[[i]]
  current_data <- extract(current_raster, SpatialPoints(coordinates, proj4string = CRS("+proj=longlat +datum=WGS84 +no_defs ")), sp = T)
  current_df <- as.data.frame(current_data)
  # Add new column with data from current raster
  climate_compiled[, paste0("raster_", i)] <- current_df[,1]
}

names(climate_compiled) <- c("longitude", "latitude", "DD5", "MAP", "MAT", "MCMT", "MWMT", "SHM")

# generate Climate PCA 
# # add identifier for each specimen 
# climate_compiled <- climate_compiled %>%
#   mutate(specimen_no = row_number())

# remove coordinates
climate_data <- climate_compiled[, -c(1, 2)]

# calculate principal components
clim_pca <- prcomp(climate_data, center = TRUE, scale = TRUE)

# # reverse signs (eigenvectors in R point in negative direction by default)
# clim_pca$rotation <- -1*clim_pca$rotation
# clim_pca$x <- -1*clim_pca$x

# display principal components
clim_pca$rotation

# Basic biplot
biplot(clim_pca, scale = 0)

# extract PCA scores for plotting
pca_scores <- as.data.frame(clim_pca$x)

# add location info
#pca_scores <- bind_cols(post_2000 %>% dplyr::select(county), pca_scores)
pca_scores_post_2000 <- bind_cols(post_2000 %>% dplyr::select(specificEpithet, year, decimalLatitude, decimalLongitude, county, institutionCode, id), pca_scores)
                        
# Calculate the variance explained by each principal component
per_variance_explained <- (clim_pca$sdev^2 / sum(clim_pca$sdev^2))*100
# Create a data frame for the variance explained
variance_df <- data.frame(
  PC = paste0("PC", 1:length(per_variance_explained)),
  Variance_Explained = per_variance_explained,
  Cumulative_Variance = cumsum(per_variance_explained)
)

## plot with ggplot
# # create PC loading vectors for plot
# loadings <- clim_pca$rotation
# # calculate arrow coordinates
# start_x <- rep(0, ncol(loadings))
# start_y <- rep(0, ncol(loadings))
# end_x <- loadings[, 1] * max(abs(clim_pca$x[, 1]))
# end_y <- loadings[, 2] * max(abs(clim_pca$x[, 2]))

## plot no loadings
# ggplot(pca_scores, aes(x=PC1, y=PC2, colour = county)) +
#   geom_point() +
#   labs(title = "Post 2000 specimens", x = "PC1 (66.6%)", y = "PC2 (24.5%)")
ggplot(pca_scores, aes(x=PC1, y=PC2, colour = county)) +
  geom_point() +
  labs(title = "Specimens from 2015-2019", x = "PC1 (72.6%)", y = "PC2 (17.6%)")
  
ggsave("Climate_PCA_2015-2019.png", width = 20, height = 10, units = "cm")

## biplot
biplot(clim_pca, cex = c(0.5,1))
ggbiplot(clim_pca)
ggsave("Climate_biplot_2015-2019.png", width = 15, height = 15, units = "cm")
```

## Interactive plot color by PC1 from above to understand climate distribution

```{r}
# import data
#CA_rad_spec <- read.csv("CCH2_all_2024.csv", header = T)

#CA_rad_spec_2015_2024 <- CA_rad_spec %>% filter(year >= 2015 & year <= 2024)

# Load the GeoTIFF file
precip_raster <- raster("MAP.tif")

# Load the California shapefile
california_shape <- st_read("./ca_state/CA_State.shp")

# Reproject raster to match the CRS of the shapefile
precip_raster_reproj <- projectRaster(precip_raster, crs = st_crs(california_shape)$proj4string)
# Crop the raster using the California shapefile
precip_raster_CA <- mask(crop(precip_raster_reproj, extent(california_shape)), california_shape)

# assign coordinate system with right syntax (this is the system the raster is in)
precip_raster_CA <- projectRaster(precip_raster_CA, crs = "+proj=longlat +datum=WGS84")

# not working and apparently not needed
# convert for leaflet
#leaflet::projectRasterForLeaflet(precip_raster_CA)

# Create a custom reversed palette function
reversed_spectral <- function(n) {
  original_palette <- brewer.pal(n, "Spectral")
  rev(original_palette)
}

reversed_PuOr <- function(n) {
  original_palette <- brewer.pal(n, "PuOr")
  rev(original_palette)
}

custom_palette <- colorRampPalette(c("#2166AC", "#4393C3", "#92C5DE", "#F4A582", "#D6604D", "#B2182B"))

pal <- colorNumeric(palette = custom_palette(100), domain = pca_scores_post_2000$PC1)

specimen_map_leaflet <- leaflet(pca_scores_post_2000 ) %>%
  addTiles() %>%  # Add OpenStreetMap tiles
  addCircleMarkers(
    lat = ~decimalLatitude, 
    lng = ~decimalLongitude, 
    radius = 7,  # Size of the markers
    color = ~pal(PC1),
    stroke = FALSE, fillOpacity = 1,  # Set fill opacity to 1 for fully opaque markers
    popup = ~paste("Scientific Name: ", specificEpithet, "<br>Year: ", year, "Herbarium:", institutionCode, "ID:", id),  # Popup with info
    label = ~paste(specificEpithet, institutionCode, id),  # add id as a label for searching
    group = "allPoints"       # Assign a generic group name
  ) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6) %>%  # Set the view over California
  addLegend(
    position = "bottomright", 
    pal = colorNumeric(reversed_spectral(9), CA_rad_spec$PC1), 
    values = ~PC1,
    title = "PC1",
    opacity = 1
  ) %>%
  # Add a search bar to find points by ID
  addSearchFeatures(
    targetGroups = "allPoints",  # Search within all markers
    options = searchFeaturesOptions(
      zoom = 10,  # Zoom level when a match is found
      openPopup = TRUE,  # Open the popup of the matching point
      textPlaceholder = "Search map..."  # Placeholder text in the search bar
    )
  )

saveWidget(specimen_map_leaflet, "interactive_CA_specimen_color_by_PC1_post_2000.html")
```

### Specimens to sample on climate PCA colored post 2000 map

```{r}
sampling <- read.csv("Raphanus_sampling_specimens.csv", header = T) 
selected <- sampling$id

specimen_map_w_sampling <- leaflet(pca_scores_post_2000) %>%
  addTiles() %>%  # Add OpenStreetMap tiles
  # Add all points
  addCircleMarkers(
    lat = ~decimalLatitude, 
    lng = ~decimalLongitude, 
    radius = 7,  # Size of the markers
    color = ~pal(PC1),
    stroke = FALSE, fillOpacity = 1,  # Set fill opacity to 1 for fully opaque markers
    popup = ~paste("Scientific Name: ", specificEpithet, "<br>Year: ", year, "Herbarium:", institutionCode, "ID:", id),  # Popup with info
    label = ~paste(specificEpithet, institutionCode, id),  # Add id as a label for searching
    group = "allPoints"       # Assign a generic group name
  ) %>%
  # Add selected points with a dark border
  addCircleMarkers(
    data = pca_scores_post_2000[pca_scores_post_2000$id %in% selected, ],
    lat = ~decimalLatitude, 
    lng = ~decimalLongitude, 
    radius = 7,  # Same size as the other points
    color = "black",  # Border color
    stroke = TRUE, weight = 3,  # Enable stroke and set border thickness
    fillColor = ~pal(PC1),  # Keep the fill color the same
    fillOpacity = 1,  # Fully opaque fill
    popup = ~paste("Scientific Name: ", specificEpithet, "<br>Year: ", year, "Herbarium:", institutionCode, "ID:", id),  # Popup with info
    label = ~paste(specificEpithet, institutionCode, id),  # Add id as a label for searching
    group = "selectedPoints"  # Separate group for selected points
  ) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6) %>%  # Set the view over California
  addLegend(
    position = "bottomright", 
    pal = colorNumeric(reversed_spectral(9), CA_rad_spec$PC1), 
    values = ~PC1,
    title = "PC1",
    opacity = 1
  ) %>%
  # Add a search bar to find points by ID
  addSearchFeatures(
    targetGroups = "allPoints",  # Search within all markers
    options = searchFeaturesOptions(
      zoom = 10,  # Zoom level when a match is found
      openPopup = TRUE,  # Open the popup of the matching point
      textPlaceholder = "Search map..."  # Placeholder text in the search bar
    )
  )

saveWidget(specimen_map_w_sampling, "bold_selected_CA_specimen_color_by_PC1_post_2000.html")

```

## Latest interactive plot of all CA specimens (09/2024)

```{r}
# import data
CA_rad_spec <- read.csv("CCH2_all_2024.csv", header = T)

specimen_map_leaflet <- leaflet(CA_rad_spec) %>%
  addTiles() %>%  # Add OpenStreetMap tiles
  # addRasterImage(precip_raster_CA,
  #                opacity = 0.7,  # Adjust opacity as needed
  #                colors = brewer.pal(9,'YlGnBu')) %>%  # Use Spectral color palette for raster
  addCircleMarkers(
    lat = ~decimalLatitude, 
    lng = ~decimalLongitude, 
    radius = 4,  # Size of the markers
    color = ~colorNumeric("Spectral", year)(year),
    stroke = FALSE, fillOpacity = 1,  # Set fill opacity to 1 for fully opaque markers
    popup = ~paste("Scientific Name: ", specificEpithet, "<br>Year: ", year, "Herbarium:", institutionCode)  # Popup with info
    label = ~paste(specificEpithet, institutionCode, id),  # add id as a label for searching
  ) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6) %>%  # Set the view over California
  addLegend(
    position = "bottomright", 
    pal = colorNumeric("Spectral", CA_rad_spec$year), 
    values = ~year,
    title = "Year",
    opacity = 1
  )

saveWidget(specimen_map_leaflet, "interactive_CA_specimen_by_year_all_CCH2.html")


###OLD CODE below

# # Convert the raster data to a data frame for use in Plotly
# precip_data <- as.data.frame(precip_raster_CA, xy = TRUE, na.rm = TRUE)
# colnames(precip_data) <- c("lon", "lat", "precip_value")
# 
# precip_map <- ggplot() + 
#   geom_raster(aes(x = lon, y = lat, fill = precip_value), data = precip_data) +
#   scale_fill_gradientn(colours = terrain.colors(10), na.value = "transparent") +
#   scale_fill_viridis_c(direction = -1)+
#   theme_void() +
#   coord_quickmap()
# 
# precip_map_plotly <- ggplotly(precip_map)
## BELOW works alone, but doesn't include climate background
# # Use plotly to make the ggplot2 map interactive
# CA_interactive_plot <- ggplotly(precip_map) %>%
#   layout(
#     mapbox = list(style = "open-street-map", zoom = 6, center = list(lat = 36.7783, lon = -119.4179)),
#     title = "California Precipitation Gradient Map"
#   )
# Interactive plot
# specimen_map_plotly <- 
#   plot_ly(data = CA_rad_spec, 
#           type = "scattermapbox", 
#           mode = "markers", 
#           hoverinfo = "text",
#           lat = ~decimalLatitude, 
#           lon = ~decimalLongitude,
#           text = ~paste("Scientific Name: ", specificEpithet, "<br>Year: ", year),
#           marker = list(size = 8, color = ~year, colorscale = "Spectral", cmin = min(CA_rad_spec$year), cmax = max(CA_rad_spec$year))) %>%
#   layout(title = "Raphanus specimen collection locations",
#          mapbox = list(style = "open-street-map", zoom = 6, center = list(lat = 36.7783, lon = -119.4179)),
#          xaxis = list(showgrid = FALSE, zeroline = FALSE),
#          yaxis = list(showgrid = FALSE, zeroline = FALSE))
#   
# # combine maps into one interactive map
# combined_plot <- subplot(
#   precip_map_plotly, 
#   specimen_map_plotly, 
#   nrows = 1, 
#   shareX = TRUE, 
#   shareY = TRUE
# ) %>% layout(
#   mapbox = list(style = "open-street-map", zoom = 6, center = list(lat = 36.7783, lon = -119.4179)),
#   title = "Combined Precipitation and Specimen Collection Map"
# )
# 
# # Render the combined interactive plot
# combined_plot
# 
# # Display the interactive plot
# CA_interactive_plot
# 
# # Save the interactive plot to share
# saveWidget(combined_plot, "interactive_CA_specimen_by_year_plot_w.precip.html")

# Attempt to add diff symbols for diff species... medium successful
# First, need to convert simplifiedName to a factor with numeric values so it can be used to code different shapes
gbif_spec2$symbol <- as.numeric(factor(gbif_spec2$simplifiedName))

# Make the plot
interactive_plot2 <- plot_geo(gbif_spec2, lat = ~decimalLatitude, lon = ~decimalLongitude, color = ~year, colors = "Spectral",
                             symbol = ~symbol, symbols = c("circle", "cross", "triangle-up", "diamond"),
                             marker = list(size = 8), mode = "markers") %>%
  layout(title = "Raphanus specimen collection locations",
         mapbox = list(style = "open-street-map", zoom = 4, center = list(lat = 37.5, lon = -95)))

# Display the interactive plot
#interactive_plot2

```

### Specimens to sample on map of all specimens

```{r}
sampling <- read.csv("Raphanus_sampling_specimens.csv", header = T) 
selected <- sampling$id
# adding bold borders to those I will sample
# Define a consistent color palette based on the full range of 'year'
colors <- colorNumeric("Spectral", domain = CA_rad_spec$year)

full_specimen_map_w_sampling <- leaflet(CA_rad_spec) %>%
  addTiles() %>%  # Add OpenStreetMap tiles
  # addRasterImage(precip_raster_CA,
  #                opacity = 0.7,  # Adjust opacity as needed
  #                colors = brewer.pal(9,'YlGnBu')) %>%  # Use Spectral color palette for raster
  addCircleMarkers(
    lat = ~decimalLatitude, 
    lng = ~decimalLongitude, 
    radius = 4,  # Size of the markers
    color = ~colors(year),
    stroke = FALSE, fillOpacity = 1,  # Set fill opacity to 1 for fully opaque markers
    popup = ~paste("Scientific Name: ", specificEpithet, "<br>Year: ", year, "Herbarium:", institutionCode, "ID:", id),  # Popup with info
    label = ~paste(specificEpithet, institutionCode, year),  # Add data as labels for searching
    group = "allPoints"
  ) %>%
  # Add selected points with a dark border
  addCircleMarkers(
    data = CA_rad_spec[CA_rad_spec$id %in% selected, ],
    lat = ~decimalLatitude, 
    lng = ~decimalLongitude, 
    radius = 4,  # Same size as the other points
    color = "black",  # Border color
    stroke = TRUE, weight = 3,  # Enable stroke and set border thickness
    fillColor = ~colors(year),  # Keep the fill color the same
    fillOpacity = 1,  # Fully opaque fill
    popup = ~paste("Scientific Name: ", specificEpithet, "<br>Year: ", year, "Herbarium:", institutionCode, "ID:", id),  # Popup with info
    label = ~paste(specificEpithet, institutionCode, year),  # Add id as a label for searching
    group = "selectedPoints"  # Separate group for selected points
  ) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6) %>%  # Set the view over California
  addLegend(
    position = "bottomright", 
    pal = colorNumeric("Spectral", CA_rad_spec$year), 
    values = ~year,
    title = "Year",
    opacity = 1
  ) %>%
  # Add a search bar to find points by ID
  addSearchFeatures(
    targetGroups = "allPoints",  # Search within all markers
    options = searchFeaturesOptions(
      zoom = 10,  # Zoom level when a match is found
      openPopup = TRUE,  # Open the popup of the matching point
      textPlaceholder = "Search map..."  # Placeholder text in the search bar
 )
  )

saveWidget(full_specimen_map_w_sampling , "bold_selected_CA_specimen_by_year_all_CCH2.html")
```

## Reading in and cleaning data for all of US

```{r import and clean data}
# Read in gbif data; search for "Raphanus L.", occurance status present, with or without coordinates, and US only.

# Citation: GBIF.org (20 July 2023) GBIF Occurrence Download https://doi.org/10.15468/dl.2srzch
#OLD -> #gbif_full <- read.csv("Raphanus_occurrences.csv", header = T)

# pare down gbif data:remove rows without a collection year, remove rows with no county AND no coordinates, remove " County" and "Co." from county names so each county only has one name.
string_to_remove <- c(" County", " Co.")

gbif <- gbif_full %>%
  select(scientificName, year, stateProvince, county, decimalLatitude, decimalLongitude, recordedBy, basisOfRecord) %>%
  drop_na(year) %>%
  filter(!is.na(decimalLatitude) & !is.na(county)) %>%
  mutate(county = str_remove_all(county, paste(string_to_remove, collapse = "|")))

# fill missing lat and long with the average lat or long for that county
# Use ifelse to conditionally fill missing lat and long only when there is a county present
gbif2 <- gbif %>%
  group_by(county) %>%
  mutate(
    decimalLatitude = ifelse(!is.na(county) & is.na(decimalLatitude), mean(decimalLatitude, na.rm = TRUE), decimalLatitude),
    decimalLongitude = ifelse(!is.na(county) & is.na(decimalLongitude), mean(decimalLongitude, na.rm = TRUE), decimalLongitude)
  ) %>%
  drop_na(decimalLatitude)

# get map
# register_google(key = "AIzaSyB39SQ16tq5W05hnFRHx6CVSKzH1iq1Mf0")
#ucsc key - AIzaSyBAl4FSPveUmJAh1g013R5mzesO7pux7g0
#map <- get_map(location = c(lon=-100 , lat= 40), zoom =3, maptype = "terrain")
# ggmap(map) # for testing map

# Separate out only the herbarium specimen records
gbif_spec2 <- gbif2 %>%
filter(basisOfRecord == "PreservedSpecimen")

# simplify scientific name to make distinguishing sativus and raphanistrum possible
# Create a new column with simplified names
gbif_spec2 <- gbif_spec2 %>%
  mutate(simplifiedName = case_when(
    startsWith(scientificName, "Raphanus raphanistrum") ~ "Raphanus raphanistrum",
    startsWith(scientificName, "Raphanus sativus") ~ "Raphanus sativus",
    TRUE ~ scientificName  # Keep the original name if it doesn't match the patterns above
  ))

```

## Plot specimen collection locations

```{r plot all specimens}

# plot all herbarium specimen locations on one map
Raph_spec_map <- ggmap(map) +
  geom_point(data = gbif_spec2, size = 2, aes(x = decimalLongitude, y = decimalLatitude, fill = year, shape = simplifiedName), color = "black") +
  ylim(25, 50) +
  xlim(-124, -65) +
  scale_fill_distiller(palette = "Spectral") +
  scale_shape_manual(values = c("Raphanus sativus" = 21, "Raphanus raphanistrum" = 24, "Raphanus L." = 23)) +
  labs(title = "Raphanus specimen collection locations")

#ggsave("gbif_all_Raphanus_specimen_map.pdf", Raph_spec_map, width = 7, height = 4.5, units = "in")

# Try making an interactive plot
interactive_plot <- 
  plot_ly(data = CA_rad_spec, 
          type = "scattermapbox", 
          mode = "markers", 
          hoverinfo = "text",
          lat = ~decimalLatitude, 
          lon = ~decimalLongitude,
          text = ~paste("Scientific Name: ", simplifiedName, "<br>Year: ", year, "Herbarium:", institutionCode),
          marker = list(size = 8, color = ~year, colorscale = "Spectral", cmin = min(gbif_spec2$year), cmax = max(gbif_spec2$year))) %>%
  layout(title = "Raphanus specimen collection locations",
         mapbox = list(style = "open-street-map", zoom = 4, center = list(lat = 37.5, lon = -95)),
         xaxis = list(showgrid = FALSE, zeroline = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE))

# Display the interactive plot
interactive_plot

# Save the interactive plot to share
saveWidget(interactive_plot, "interactive_specimen_by_year_plot.html")

# Attempt to add diff symbols for diff species... medium successful
# First, need to convert simplifiedName to a factor with numeric values so it can be used to code different shapes
gbif_spec2$symbol <- as.numeric(factor(gbif_spec2$simplifiedName))

# Make the plot
interactive_plot2 <- plot_geo(gbif_spec2, lat = ~decimalLatitude, lon = ~decimalLongitude, color = ~year, colors = "Spectral",
                             symbol = ~symbol, symbols = c("circle", "cross", "triangle-up", "diamond"),
                             marker = list(size = 8), mode = "markers") %>%
  layout(title = "Raphanus specimen collection locations",
         mapbox = list(style = "open-street-map", zoom = 4, center = list(lat = 37.5, lon = -95)))

# Display the interactive plot
#interactive_plot2

# Plot only raphanistrum
gbif_spec_raph <- gbif_spec2 %>%
filter(scientificName == "Raphanus raphanistrum L.")
 
Raph_spec_raph_map <- ggmap(map) +
  geom_point(data = gbif_spec_raph, size = 2, pch = 21, colour="black", aes(x = decimalLongitude, y = decimalLatitude, fill = year)) +
  ylim(25,50) +
  xlim(-124, -65) +
  scale_fill_distiller(palette = "Spectral") +
  labs(title = "gbif R. raphinastrum specimen collection locations")

# only sativus
gbif_spec_sati <- gbif_spec2 %>%
filter(scientificName == "Raphanus sativus L.")
 
Raph_spec_sati_map <- ggmap(map) +
  geom_point(data = gbif_spec_sati, size = 2, pch = 21, colour="black", aes(x = decimalLongitude, y = decimalLatitude, fill = year)) +
  ylim(25,50) +
  xlim(-124, -65) +
  scale_fill_distiller(palette = "Spectral") +
  labs(title = "gbif R. sativus specimen collection locations")

# only hybrids - not bothering, it is only 2 specimens from CA in 2007 and 2013
gbif_spec_hyb <- gbif_spec2 %>%
filter(scientificName == "Raphanus raphanistrum × sativus")

# no genus and hybrids - not bothering, only 35, quite a few in mexico
gbif_spec_hyb.no <- gbif_spec2 %>%
filter(scientificName %in% c("","Raphanus raphanistrum × sativus"))
 
# patchwork multipanel plot to compare species
gbif_2panel <- Raph_spec_sati_map/Raph_spec_raph_map

ggsave("gbif_R.sativus_R.raphinastrum_specimens_map.pdf", gbif_2panel, width = 7, height = 9, units = "in")
```

## California only specimen locations

```{r}
# Load map of CA (FIX BOX)
map <- get_map(location = c(lon=-119 , lat= 37.5), zoom =6, maptype = "terrain")

#PICK UP HERE: make 2 CA plots: pre 1940, 1940-1980, post 1980
# code to MODIFY below

# Plot only pre 1940 specimen locations
gbif_spec_pre1940 <- gbif_spec2 %>% 
  filter(year <= 1940)

Raph_spec_map_pre1940 <- ggmap(map) +
  geom_point(data = gbif_spec_pre1940, size = 2, colour="black", aes(x = decimalLongitude, y = decimalLatitude, fill = year, shape = simplifiedName)) +
  ylim(32.8,41.7) +
  xlim(-124.1, -114.5) +
  scale_fill_distiller(palette = "Spectral") +
  scale_shape_manual('Purported species', values = c("Raphanus sativus" = 21, "Raphanus raphanistrum" = 24), labels = c("weed", "crop")) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  guides(fill = guide_colourbar(reverse = TRUE))

# Plot 1940-1980 specimen locations
gbif_spec_1940to1980 <- gbif_spec2 %>% 
  filter(year <= 1980 & year >= 1940)

Raph_spec_map_1940to1980 <- ggmap(map) +
  geom_point(data = gbif_spec_1940to1980, size = 2, colour="black", aes(x = decimalLongitude, y = decimalLatitude, fill = year, shape = simplifiedName)) +
  ylim(32.8,41.7) +
  xlim(-124.1, -114.5) +
  scale_fill_distiller(palette = "Spectral") +
  scale_shape_manual('Purported species', values = c("Raphanus sativus" = 21, "Raphanus raphanistrum" = 24), labels = c("weed", "crop")) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  guides(fill = guide_colourbar(reverse = TRUE))

# Plot post 1980 specimen locations
gbif_spec_post1980 <- gbif_spec2 %>% 
  filter(year >= 1980)

Raph_spec_map_post1980 <- ggmap(map) +
  geom_point(data = gbif_spec_post1980, size = 2, colour="black", aes(x = decimalLongitude, y = decimalLatitude, fill = year, shape = simplifiedName)) +
  ylim(32.8,41.7) +
  xlim(-124.1, -114.5) +
  scale_fill_distiller(palette = "Spectral") +
  scale_shape_manual('Purported species', values = c("Raphanus sativus" = 21, "Raphanus raphanistrum" = 24), labels = c("weed", "crop")) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  guides(fill = guide_colourbar(reverse = TRUE))

# plot in a multipanel 
three_panel_CA_Raph_map <- Raph_spec_map_pre1940 / Raph_spec_map_1940to1980 / Raph_spec_map_post1980

# save plot
ggsave("three_panel_CA_Raph_map.pdf", three_panel_CA_Raph_map, width = 4.5, height = 12, units = "in")

```

## Plot and count by county

```{r plot by county}

# plot by county in California
CA_barplot_by_spp.co <- gbif_spec2 %>% 
  filter(stateProvince == "California") %>% 
  group_by(county, simplifiedName) %>%
  summarise(count = n()) %>% 
  ggplot(aes(x=county, y=count, fill=simplifiedName)) + 
  geom_bar(stat = "identity") +
  labs(y = "Number of Individuals") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggsave("CA_barplot_by_spp.co.pdf", CA_barplot_by_spp.co, width = 14, height = 7, units = "in")

# Get just the per county counts from most to least:
CA_per_co_counts <- gbif_spec2 %>% 
  filter(stateProvince == "California") %>% 
  group_by(county, simplifiedName) %>%
  summarise(count = n()) %>% 
  arrange(desc(count))

write.csv(CA_per_co_counts, file = "CA_per_co_counts.csv", row.names = FALSE)

# plot by county in NW
NW_barplot_by_spp.co <- gbif_spec2 %>% 
  filter(stateProvince %in% c("Washington","Oregon")) %>% 
  group_by(county, simplifiedName) %>%
  summarise(count = n()) %>% 
  ggplot(aes(x=county, y=count, fill=simplifiedName)) + 
  geom_bar(stat = "identity") +
  labs(y = "Number of Individuals") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggsave("NW_barplot_by_spp.co.pdf", NW_barplot_by_spp.co, width = 14, height = 7, units = "in")

# Get just the per county counts from most to least:
NW_per_co_counts <- gbif_spec2 %>% 
  filter(stateProvince %in% c("Washington","Oregon")) %>% 
  group_by(county, simplifiedName) %>%
  summarise(count = n()) %>% 
  arrange(desc(count))

write.csv(NW_per_co_counts, file = "NW_per_co_counts.csv", row.names = FALSE)

# plot by county in NE
NE_barplot_by_spp.co <- gbif_spec2 %>% 
  filter(stateProvince %in% c("Maine", "Vermont", "New Hampshire", "Pennsylvania", "Massachusetts" , "Rhode Islande", "Connecticut", "New York", "New Jersey","Delaware", "Maryland", "District of Columbia")) %>% 
  group_by(county, simplifiedName) %>%
  summarise(count = n()) %>% 
  ggplot(aes(x=county, y=count, fill=simplifiedName)) + 
  geom_bar(stat = "identity") +
  labs(y = "Number of Individuals") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggsave("NE_barplot_by_spp.co.pdf", NE_barplot_by_spp.co, width = 14, height = 7, units = "in")

# Get just the per county counts from most to least:
NE_per_co_counts <- gbif_spec2 %>% 
  filter(stateProvince %in% c("Maine", "Vermont", "New Hampshire", "Pennsylvania", "Massachusetts" , "Rhode Islande", "Connecticut", "New York", "New Jersey","Delaware", "Maryland", "District of Columbia")) %>% 
  group_by(county, simplifiedName) %>%
  summarise(count = n()) %>% 
  arrange(desc(count))

write.csv(NE_per_co_counts, file = "NE_per_co_counts.csv", row.names = FALSE)

# plot by county in SE
SE_barplot_by_spp.co <- gbif_spec2 %>% 
  filter(stateProvince %in% c("Virginia", "North Carolina", "South Carolina", "Tennesee", "Georgia", "Alabama", "Florida", "Mississippi", "Louisiana", "Arkansas")) %>% 
  group_by(county, simplifiedName) %>%
  summarise(count = n()) %>% 
  ggplot(aes(x=county, y=count, fill=simplifiedName)) + 
  geom_bar(stat = "identity") +
  labs(y = "Number of Individuals") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggsave("SE_barplot_by_spp.co.pdf", SE_barplot_by_spp.co, width = 14, height = 7, units = "in")

# Get just the per county counts from most to least:
SE_per_co_counts <- gbif_spec2 %>% 
  filter(stateProvince %in% c("Virginia", "North Carolina", "South Carolina", "Tennesee", "Georgia", "Alabama", "Florida", "Mississippi", "Louisiana", "Arkansas")) %>% 
  group_by(county, simplifiedName) %>%
  summarise(count = n()) %>% 
  arrange(desc(count))

write.csv(SE_per_co_counts, file = "SE_per_co_counts.csv", row.names = FALSE)
```

## Plotting only old specimen locations

```{r plot based on time period}
## Plot only old specimens
# Plot only pre 1900 specimen locations
gbif_spec_pre1900 <- gbif_spec2 %>% 
  filter(year <= 1900)

pre1900_Raph_spec_map <- ggmap(map) +
  geom_point(data = gbif_spec_pre1900, size = 2, colour="black", aes(x = decimalLongitude, y = decimalLatitude, fill = year, shape = simplifiedName)) +
  ylim(25,50) +
  xlim(-124, -65) +
  scale_fill_distiller(palette = "Spectral") +
  scale_shape_manual(values = c("Raphanus sativus" = 21, "Raphanus raphanistrum" = 24, "Raphanus L." = 23)) +
  labs(title = "1900 and older Raphanus specimen collection locations")

# Plot 1900 to 1920 specimen locations
gbif_spec_1900to1920 <- gbif_spec2 %>% 
  filter(year <= 1920 & year >= 1900)

btw_1900.1920_Raph_spec_map <- ggmap(map) +
  geom_point(data = gbif_spec_1900to1920, size = 2, colour="black", aes(x = decimalLongitude, y = decimalLatitude, fill = year, shape = simplifiedName)) +
  ylim(25,50) +
  xlim(-124, -65) +
  scale_fill_distiller(palette = "Spectral") +
  scale_shape_manual(values = c("Raphanus sativus" = 21, "Raphanus raphanistrum" = 24, "Raphanus L." = 23)) +
  labs(title = "1900-1920 Raphanus specimen collection locations")

# save multipanel 
old_spec <- pre1900_Raph_spec_map/btw_1900.1920_Raph_spec_map

ggsave("pre_1900+1900-1920_specimen_maps.pdf", old_spec, width = 7, height = 9, units = "in")

# create plots for each 20 year period to inform sampling plan

# 1920-40
gbif_spec_1920to1940 <- gbif_spec2 %>% 
  filter(year >= 1920 & year <= 1940)

btw_1920.1940_Raph_spec_map <- ggmap(map) +
  geom_point(data = gbif_spec_1920to1940, size = 2, colour="black", aes(x = decimalLongitude, y = decimalLatitude, fill = year, shape = simplifiedName)) +
  ylim(25,50) +
  xlim(-124, -65) +
  scale_fill_distiller(palette = "Spectral") +
  scale_shape_manual(values = c("Raphanus sativus" = 21, "Raphanus raphanistrum" = 24, "Raphanus L." = 23)) +
  labs(title = "1920-1940 Raphanus specimen collection locations")

# 1940-60
gbif_spec_1940to1960 <- gbif_spec2 %>% 
  filter(year >= 1940 & year <= 1960)

btw_1940.1960_Raph_spec_map <- ggmap(map) +
  geom_point(data = gbif_spec_1940to1960, size = 2, colour="black", aes(x = decimalLongitude, y = decimalLatitude, fill = year, shape = simplifiedName)) +
  ylim(25,50) +
  xlim(-124, -65) +
  scale_fill_distiller(palette = "Spectral") +
  scale_shape_manual(values = c("Raphanus sativus" = 21, "Raphanus raphanistrum" = 24, "Raphanus L." = 23)) +
  labs(title = "1940-1960 Raphanus specimen collection locations")

# 1960-80
gbif_spec_1960to1980 <- gbif_spec2 %>% 
  filter(year >= 1960 & year <= 1980)

btw_1960.1980_Raph_spec_map <- ggmap(map) +
  geom_point(data = gbif_spec_1960to1980, size = 2, colour="black", aes(x = decimalLongitude, y = decimalLatitude, fill = year, shape = simplifiedName)) +
  ylim(25,50) +
  xlim(-124, -65) +
  scale_fill_distiller(palette = "Spectral") +
  scale_shape_manual(values = c("Raphanus sativus" = 21, "Raphanus raphanistrum" = 24, "Raphanus L." = 23)) +
  labs(title = "1960-1980 Raphanus specimen collection locations")

# 1980-2000
gbif_spec_1980to2000 <- gbif_spec2 %>% 
  filter(year >= 1980 & year <= 2000)

btw_1980.2000_Raph_spec_map <- ggmap(map) +
  geom_point(data = gbif_spec_1980to2000, size = 2, colour="black", aes(x = decimalLongitude, y = decimalLatitude, fill = year, shape = simplifiedName)) +
  ylim(25,50) +
  xlim(-124, -65) +
  scale_fill_distiller(palette = "Spectral") +
  scale_shape_manual(values = c("Raphanus sativus" = 21, "Raphanus raphanistrum" = 24, "Raphanus L." = 23)) +
  labs(title = "1980-2000 Raphanus specimen collection locations")

# 2000-2023
gbif_spec_2000to2023 <- gbif_spec2 %>% 
  filter(year >= 2000)

btw_2000.2023_Raph_spec_map <- ggmap(map) +
  geom_point(data = gbif_spec_2000to2023, size = 2, colour="black", aes(x = decimalLongitude, y = decimalLatitude, fill = year, shape = simplifiedName)) +
  ylim(25,50) +
  xlim(-124, -65) +
  scale_fill_distiller(palette = "Spectral") +
  scale_shape_manual(values = c("Raphanus sativus" = 21, "Raphanus raphanistrum" = 24, "Raphanus L." = 23)) +
  labs(title = "2000-2023 Raphanus specimen collection location")
  
# make into a multipanel plot
map_panel_20yr_periods <- (btw_1900.1920_Raph_spec_map/btw_1920.1940_Raph_spec_map/btw_1940.1960_Raph_spec_map) | (btw_1960.1980_Raph_spec_map/btw_1980.2000_Raph_spec_map/btw_2000.2023_Raph_spec_map)

# save
ggsave("map_panel_20yr_periods.pdf", map_panel_20yr_periods, width = 14, height = 15, units = "in")

```

# Plot all occurance/observation data

```{r}
# plot all points (including observation data) on map
Raph_all_map <- ggmap(map) +
  geom_point(data = gbif, size = 2, pch = 21, colour="black", aes(x = decimalLongitude, y = decimalLatitude, fill = year)) +
  ylim(14,52) +
  xlim(-127, -56) +
  scale_fill_distiller(palette = "Spectral") +
  labs(title = "gbif all Raphanus spp. records")

## plot all occurance/observation data
# all spp together
gbif_occ <- gbif %>%
filter(basisOfRecord %in% c("HUMAN_OBSERVATION", "OCCURRENCE", "OBSERVATION"))
 
Raph_occ_map <- ggmap(map) +
  geom_point(data = gbif_occ, size = 2, pch = 21, colour="black", aes(x = decimalLongitude, y = decimalLatitude, fill = year)) +
  ylim(14,52) +
  xlim(-127, -56) +
  scale_fill_distiller(palette = "Spectral") +
  labs(title = "gbif Raphanus spp. occurance/observation locations")

# only raphinastrum

# only sativus

#ggsave("gbif_Raphanus_occurance_map.pdf", Raph_occ_map)

```

## Tables and plots of number of specimens through time in different regions

```{r}
## Make a table for each 20 year period since 1920 (oldest is 1927)
# the dataframe has already been filtered to only include herbarium specimen records from North America
# Create a new column for 20-year periods
gbif_spec2 <- gbif_spec2 %>%
  mutate(period = paste0(floor(year / 20) * 20, "-", floor(year / 20) * 20 + 19))

# Count the number of rows for each 20-year period
count_by_period_table <- gbif_spec2 %>%
  group_by(period) %>%
  summarise(count = n())

## Now I will break it down by focal region:

# California
# Count the number of rows for each 20-year period, each county, and each simplifiedName
California_count_by_per.co.sp <- gbif_spec2 %>%
  filter(stateProvince == "California") %>%
  group_by(period, stateProvince, county, simplifiedName) %>%
  summarise(count = n())

# export to csv
write.csv(California_count_by_per.co.sp, file = "California_specimen_count_by_per.co.sp.csv", row.names = FALSE)

# Plot just San Diego because the most samples 
SD_barplot_by_spp.period <- California_count_by_per.co.sp %>% 
  filter(county == "San Diego") %>% 
  ggplot(aes(x=period, y=count, fill=simplifiedName)) + 
  geom_bar(stat = "identity") +
  labs(y = "Number of Individuals") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggsave("SD_barplot_by_spp.period.pdf", SD_barplot_by_spp.period, width = 14, height = 7, units = "in")


# Washinton & Oregon (NW)
# Count the number of rows for each 20-year period, each county, and each simplifiedName
NW_count_by_per.co.sp <- gbif_spec2 %>%
  filter(stateProvince %in% c("Washington","Oregon")) %>%
  group_by(period, stateProvince, county, simplifiedName) %>%
  summarise(count = n())

# export to csv
write.csv(NW_count_by_per.co.sp, file = "NW_specimen_count_by_per.co.sp.csv", row.names = FALSE)

# The North East (NE)
# Count the number of rows for each 20-year period, each county, and each simplifiedName
NE_count_by_per.co.sp <- gbif_spec2 %>%
  filter(stateProvince %in% c("Maine", "Vermont", "New Hampshire", "Pennsylvania", "Massachusetts" , "Rhode Islande", "Connecticut", "New York", "New Jersey","Delaware", "Maryland", "District of Columbia")) %>%
  group_by(period, stateProvince, county, simplifiedName) %>%
  summarise(count = n())

# export to csv
write.csv(NE_count_by_per.co.sp, file = "NE_specimen_count_by_per.co.sp.csv", row.names = FALSE)

# The South East (SE)
# Count the number of rows for each 20-year period, each county, and each simplifiedName
SE_count_by_per.co.sp <- gbif_spec2 %>%
  filter(stateProvince %in% c("Virginia", "North Carolina", "South Carolina", "Tennesee", "Georgia", "Alabama", "Florida", "Mississippi", "Louisiana", "Arkansas")) %>%
  group_by(period, stateProvince, county, simplifiedName) %>%
  summarise(count = n())

# export to csv
write.csv(SE_count_by_per.co.sp, file = "SE_specimen_count_by_per.co.sp.csv", row.names = FALSE)
```

# Plot 5 year intervals

```{r plotting 5 yr intervals}
# Create a new column for 5-year periods
gbif_spec2 <- gbif_spec2 %>%
  mutate(period5 = paste0(floor(year / 5) * 5, "-", floor(year / 5) * 5 + 4))

CA_rad_spec <- CA_rad_spec  %>%
  mutate(period5 = paste0(floor(year / 5) * 5, "-", floor(year / 5) * 5 + 4))

# Count the number of rows for each 5-year period
count_by_period5_table <- gbif_spec2 %>%
  group_by(period5, simplifiedName) %>%
  summarise(count = n())

count_by_period5_table <- CA_rad_spec %>%
  group_by(period5, specificEpithet) %>%
  summarise(count = n())

# Plot frequency through time in 5 yr intervals, color by spp, all data
barplot_by_spp.period5 <- count_by_period5_table %>% 
  ggplot(aes(x=period5, y=count, fill=specificEpithet)) + 
  geom_bar(stat = "identity") +
  labs(y = "Number of Individuals") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggsave("all_barplot_by_spp.period5.pdf", barplot_by_spp.period5, width = 14, height = 7, units = "in")

## Now I will break it down by focal region:

#SoCA
SoCA_count_by_per5.sp <- gbif_spec2 %>%
  filter(stateProvince == "California" & county %in% c("San Diego", "Santa Barbara", "Orange", "Riverside", "Los Angeles", "Ventura", "San Bernardino", "San Luis Obispo")) %>%
  group_by(period5, simplifiedName) %>%
  summarise(count = n())

# Plot frequency through time in 5 yr intervals, color by spp, all data
SoCA_barplot_by_spp.period5 <- SoCA_count_by_per5.sp %>% 
  ggplot(aes(x=period5, y=count, fill=simplifiedName)) + 
  geom_bar(stat = "identity") +
  labs(y = "Number of Individuals") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggsave("SoCA_barplot_by_spp.period5.pdf", SoCA_barplot_by_spp.period5, width = 14, height = 7, units = "in")

# look at yearly instead of by period:
SoCA_yrly_count_by_sp <- gbif_spec2 %>%
  filter(stateProvince == "California" & county %in% c("San Diego", "Santa Barbara", "Orange", "Riverside", "Los Angeles", "Ventura", "San Bernardino", "San Luis Obispo")) %>%
  group_by(year, simplifiedName) %>%
  summarise(count = n())

# Plot frequency through time in 5 yr intervals, color by spp, all data
SoCA_yrly_barplot_by_spp <- SoCA_yrly_count_by_sp %>% 
  ggplot(aes(x=year, y=count, fill=simplifiedName)) + 
  geom_bar(stat = "identity") +
  labs(y = "Number of Individuals") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggsave("SoCA_yrly_barplot_by_spp.pdf", SoCA_yrly_barplot_by_spp, width = 14, height = 7, units = "in")

SoCA_1965 <- gbif_spec2 %>% filter(stateProvince == "California" & county %in% c("San Diego", "Santa Barbara", "Orange", "Riverside", "Los Angeles", "Ventura", "San Bernardino", "San Luis Obispo") & year =="1965") 

# looking at crazy high collection count in soCal in 1965 in more detail
ggmap(map) +
  geom_point(data=SoCA_1965, size = 2, aes(x = decimalLongitude, y = decimalLatitude, shape = simplifiedName), color = "black") +
  ylim(30, 35) +
  xlim(-122, -115) +
  scale_shape_manual(values = c("Raphanus sativus" = 21, "Raphanus raphanistrum" = 24, "Raphanus L." = 23)) +
  labs(title = "Raphanus specimen collection locations")

```

#Playing with some summary plots

```{r}
# Stacked Bar Plot - too much with all the counties, but will be helpful looking at chosen counties and shorter time intervals
ggplot(California_count_by_per.co.sp, aes(x = period, y = count, fill = simplifiedName)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ county, scales = "free") +
  labs(title = "Specimen Counts by Scientific Name for each County and Time Period",
       x = "Time Period",
       y = "Count of Specimens") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

# CCH2 data only

```{r}
# Read in occurance data from CCH2 search for "Raphanus" (had to delete verbatim coordinates because contain crazy characters)
occ_full <- read.csv("Raphanus_occurrences.csv", header = T)

# pare down data: keep only reccords from CA, remove rows without a collection year, remove " County" and "Co." from county names so each county only has one name. 
string_to_remove <- c(" County", " Co.")
occ <- occ_full %>% 
  select(id, simplifiedName, genus, specificEpithet, year, stateProvince, county, decimalLatitude, decimalLongitude) %>% 
  filter(stateProvince == "California") %>% 
  drop_na(year, county) %>% 
  mutate(county = str_remove_all(county, paste(string_to_remove, collapse = "|")))
  
# fill in missing coordinates with a genereic per county point
# determine which counties are represented:
unique(occ$county)

# fill missing lat and long with the average lat or long for that county
occ2 <- occ %>%
  filter(!county =="") %>% 
  group_by(county) %>%
  mutate(decimalLatitude = ifelse(is.na(decimalLatitude), mean(decimalLatitude, na.rm = TRUE), decimalLatitude),
         decimalLongitude = ifelse(is.na(decimalLongitude), mean(decimalLongitude, na.rm = TRUE), decimalLongitude)) %>% 
  drop_na(decimalLatitude)
  
## Plot on map of CA  
# compute the bounding box for sample map
bbox <- make_bbox(lat = decimalLatitude, lon = decimalLongitude, data = occ2)

# adjust right bound to include all of CA
bbox["right"] <- -114.00

# get map of CA
# register_google(key = "AIzaSyB39SQ16tq5W05hnFRHx6CVSKzH1iq1Mf0")
california_map <- get_map(location = bbox, zoom = 6, maptype = "terrain")

# reverse order for plotting
occ2 <- occ2[order(-occ2$year), ]

# plot points on map
Raph_map <- ggmap(california_map) +
  geom_point(data = occ2, size = 2, pch = 21, colour="black", aes(x = decimalLongitude, y = decimalLatitude, fill = year)) +
  scale_fill_distiller(palette = "Spectral") +
  labs(title = "CCH2 all Raphanus spp. specimen locations")
  
ggsave("CCH2_Raphanus_specimen_map.pdf", Raph_map)

## OG simple map - gray CA outline
# # Get map data for California
# ca_map <- map_data("state", "california")
# 
# # Create ggplot object with California outline
# ggplot(ca_map, aes(x = long, y = lat, group = group)) +
#   geom_polygon(fill = "lightgray", colour = "white") +
#   coord_fixed(1.3)  # Adjust the aspect ratio if needed
# 
# # Add points to the plot
# ggplot(ca_map, aes(x = long, y = lat)) +
#   geom_polygon(fill = "lightgray", colour = "white") +
#   geom_point(data = occ2, aes(x = decimalLongitude, y = decimalLatitude, color = year)) +
#   scale_color_gradient(low = "blue", high = "red", limits = c(1857, 2023)) +
#   labs(title = "Lat Long Coordinates in California by Year", x = "Longitude", y = "Latitude") +
#   coord_fixed(1.3)  # Adjust the aspect ratio if needed

```

No need to combine as done below. gbif includes all CCH records...

Combine CCH2 and gbif specimen record data and plot again

```{r}

# Read in occurance data from CCH2 search for "Raphanus" (had to delete verbatim coordinates because contain crazy characters)

cch_full <- read.csv("CCH2_Raphanus_occurrences.csv", header = T)

# pare down CCH2 data: remove rows without a collection year, remove rows with no county AND no coordinates, remove " County" and "Co." from county names so each county only has one name.

string_to_remove <- c(" County", " Co.")

cch <- cch_full %>%

  select(simplifiedName, year, stateProvince, county, decimalLatitude, decimalLongitude, recordedBy, basisOfRecord) %>%

#  filter(stateProvince == "California") %>%

  drop_na(year) %>%

  filter(!is.na(decimalLatitude) & !is.na(county)) %>%

  mutate(county = str_remove_all(county, paste(string_to_remove, collapse = "|")))

# Read in gbif data; search for "Raphanus L.", occurance status present, with or without coordinates, and US only.

# Citation: GBIF.org (20 July 2023) GBIF Occurrence Download https://doi.org/10.15468/dl.2srzch

gbif_full <- read.csv("gbif_Raphanus_all_present.csv", header = T)

# pare down gbif data:

gbif <- gbif_full %>%

  select(simplifiedName, year, stateProvince, county, decimalLatitude, decimalLongitude, recordedBy, basisOfRecord) %>%

  drop_na(year) %>%

  filter(!is.na(decimalLatitude) & !is.na(county)) %>%

  mutate(county = str_remove_all(county, paste(string_to_remove, collapse = "|")))

# Join the datasets

comb_full <- gbif %>%

  full_join(cch, by = c("scientificName", "year", "stateProvince", "county", "decimalLatitude", "decimalLongitude", "recordedBy", "basisOfRecord"))

comb_anti <- gbif %>%

  anti_join(cch, by = c("scientificName", "year", "stateProvince", "county", "decimalLatitude", "decimalLongitude", "recordedBy", "basisOfRecord"))

#Test joining with mapping:

# get map

# register_google(key = "AIzaSyB39SQ16tq5W05hnFRHx6CVSKzH1iq1Mf0")

map <- get_map(location = c(lon=-100 , lat= 40), zoom =3, maptype = "terrain")

ggmap(map)

comb_spec <- comb_full %>%

filter(basisOfRecord == "PRESERVED_SPECIMEN")

Raph_spec_map_full <- ggmap(map) +

  geom_point(data = comb_spec, size = 2, pch = 21, colour="black", aes(x = decimalLongitude, y = decimalLatitude, fill = year)) +

  ylim(14,52) +

  xlim(-127, -56) +

  scale_fill_distiller(palette = "Spectral") +

  labs(title = "Raphanus spp. specimen collection locations")

# fill missing lat and long with the average lat or long for that county

comb2 <- comb%>%

  filter(!county =="") %>%

  group_by(county) %>%

  mutate(decimalLatitude = ifelse(is.na(decimalLatitude), mean(decimalLatitude, na.rm = TRUE), decimalLatitude),

         decimalLongitude = ifelse(is.na(decimalLongitude), mean(decimalLongitude, na.rm = TRUE), decimalLongitude)) %>%

  drop_na(decimalLatitude)

```
